---
title: "Untitled"
output: html_document
date: "2025-11-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(tidyverse)
library(here)

```

```{r} 


ew <- read_csv(here("data", "ew_data_4.csv"))

ew_filtered <- ew %>%
  rename(
    ethnic_group    = `Ethnic group (6 categories) Code`,
    degree          = `Highest level of qualification (7 categories) Code`,
    age             = `Age (12 categories) Code`,
    sex             = `Sex (2 categories) Code`,
    regions         = Regions, 
    dependent_child = `Dependent children in household and their age (6 categories) Code`,
    count           = Observation
  ) %>%
  filter(
    !age %in% c(1, 2, 9, 10, 11,12),              
    dependent_child %in% c(1,2),
    sex == 1  
  ) %>%
  mutate(
    ethnicity = case_when(
      ethnic_group == 1 ~ "Asian",
      ethnic_group == 2 ~ "Black",
      ethnic_group == 3 ~ "Mixed",  
      ethnic_group == 4 ~ "White", 
      ethnic_group == 5 ~ "Other",  
      TRUE ~ "Unknown"
    ),
    age_group = case_when(
      age %in% c(3, 4) ~ "19-29",
      age %in% c(5, 6) ~ "30-39",
      age %in% c(7, 8) ~ "40+",
      TRUE ~ "Unknown"
    ),
    degree_category = case_when(
      degree %in% c(-8, 0, 1, 2, 3, 5) ~ "No degree",
      degree == 4 ~ "Degree",
      TRUE ~ "Unknown"
    )
  )



ew_eth_deg_counts <- ew_filtered %>%
  group_by(ethnicity, degree_category) %>%
  summarise(total_count = sum(count, na.rm = TRUE)) %>%
  ungroup()


ew_age_counts <- ew_filtered %>%
  group_by(age_group) %>%
  summarise(total_count = sum(count, na.rm = TRUE)) %>%
  ungroup()


region_counts <- ew_filtered %>%
  group_by(regions) %>%
  summarise(total_count = sum(count, na.rm = TRUE)) %>%
  ungroup()

```

```{r}

scotland <- read_csv(here("data", "scot_sex_age_eth_ed_depend.csv"))

scotland_filter <- scotland  %>% 
  fill(sex, age_group, ethnicity,degree, .direction="down") %>% 
  filter(sex == "Female") %>% 
  filter(grepl("With dependent", household)) %>% 
  filter(age_group %in% c("Aged 18 to 19", "Aged 20 to 24", "Aged 25 to 29", "Aged 30 to 34", "Aged 35 to 39", "Aged 40 to 44")) %>% 
  mutate(degree = case_when(
    grepl("^Degree", degree) ~ "Has Degree", 
    grepl("Not applicable", degree) ~ NA_character_, 
    TRUE ~ "No Degree"
  ), 
  ethnicity = case_when(
    grepl("African|Caribbean", ethnicity) ~ "Black", 
    grepl("Asian", ethnicity) ~ "Asian", 
    grepl("White", ethnicity) ~ "White", 
    grepl("Mixed", ethnicity) ~ "Mixed", 
    grepl("Other", ethnicity) ~ "Other", 
    TRUE ~ NA_character_
  )) %>% 
  group_by(ethnicity, degree) %>% 
  summarise(count = sum(count)) %>% 
  ungroup() %>% 
  mutate(pct = count/sum(count))

```

```{r}

library(tidyverse)
library(here)



# First, fill down the NA values (sparse format issue)
scotland <- scotland %>%
  fill(sex, age_group, ethnicity, degree, .direction = "down")

# Clean and recode Scotland data to match England format
scotland_filtered <- scotland %>%
  filter(
    sex == "Female",
    grepl("With dependent", household),
    age_group %in% c("Aged 18 to 19", "Aged 20 to 24", "Aged 25 to 29", 
                     "Aged 30 to 34", "Aged 35 to 39", "Aged 40 to 44")
  ) %>%
  # Recode degree
  mutate(
    degree_category = case_when(
      grepl("^Degree", degree) ~ "Degree",
      grepl("Not applicable", degree) ~ NA_character_,
      TRUE ~ "No degree"
    )
  ) %>%
  # Remove rows with NA degree
  filter(!is.na(degree_category)) %>%
  # Recode ethnicity
  mutate(
    ethnicity = case_when(
      grepl("African|Caribbean", ethnicity) ~ "Black",
      grepl("Asian", ethnicity) ~ "Asian",
      grepl("White", ethnicity) ~ "White",
      grepl("Mixed", ethnicity) ~ "Mixed",
      grepl("Other", ethnicity) ~ "Other",
      TRUE ~ NA_character_
    )
  ) %>%
  # Remove rows with NA ethnicity
  filter(!is.na(ethnicity)) %>%
  # Recode age groups to match England (19-29, 30-39, 40+)
  mutate(
    age_group = case_when(
      age_group %in% c("Aged 18 to 19", "Aged 20 to 24", "Aged 25 to 29") ~ "19-29",
      age_group %in% c("Aged 30 to 34", "Aged 35 to 39") ~ "30-39",
      age_group %in% c("Aged 40 to 44") ~ "40+",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(age_group)) %>%
  # Add region
  mutate(regions = "Scotland")

# Create summary tables
scotland_eth_deg_counts <- scotland_filtered %>%
  group_by(ethnicity, degree_category) %>%
  summarise(total_count = sum(count, na.rm = TRUE), .groups = "drop")

scotland_age_counts <- scotland_filtered %>%
  group_by(age_group) %>%
  summarise(total_count = sum(count, na.rm = TRUE), .groups = "drop")

scotland_region_counts <- scotland_filtered %>%
  group_by(regions) %>%
  summarise(total_count = sum(count, na.rm = TRUE), .groups = "drop")




combined_data <- bind_rows(
  ew_filtered %>% 
    select(ethnicity, degree_category, age_group, regions, count) %>%
    mutate(country = "England & Wales"),
  scotland_filtered %>% 
    select(ethnicity, degree_category, age_group, regions, count) %>%
    mutate(country = "Scotland")
)



# Combined summary tables
combined_eth_deg_counts <- combined_data %>%
  group_by(ethnicity, degree_category) %>%
  summarise(total_count = sum(count, na.rm = TRUE), .groups = "drop")

combined_age_counts <- combined_data %>%
  group_by(age_group) %>%
  summarise(total_count = sum(count, na.rm = TRUE), .groups = "drop")

combined_region_counts <- combined_data %>%
  group_by(regions) %>%
  summarise(total_count = sum(count, na.rm = TRUE), .groups = "drop")




# Table 1: Overall ethnicity x degree counts and percentages
combined_eth_deg_counts <- combined_data %>%
  group_by(ethnicity, degree_category) %>%
  summarise(total_count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  mutate(pct = round(total_count / sum(total_count) * 100, 2))

# Table 2: Within-ethnicity degree attainment rates
combined_eth_deg_within <- combined_data %>%
  group_by(ethnicity, degree_category) %>%
  summarise(total_count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  group_by(ethnicity) %>%
  mutate(pct_within_ethnicity = round(total_count / sum(total_count) * 100, 2)) %>%
  ungroup()

combined_age_counts <- combined_data %>%
  group_by(age_group) %>%
  summarise(total_count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  mutate(pct = round(total_count / sum(total_count) * 100, 2))

combined_region_counts <- combined_data %>%
  group_by(regions) %>%
  summarise(total_count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  mutate(pct = round(total_count / sum(total_count) * 100, 2))




# Save ouput 

write_csv(combined_age_counts, here("output", "age.csv"))
write_csv(combined_eth_deg_counts, here("output", "eth_deg_counts.csv"))
write.csv(combined_eth_deg_within, here("output", "eth_deg_within.csv"))
write_csv(combined_region_counts, here("output", "region_counts.csv"))




